\vspace{-1cm} Using Bayesian non-linear models to uncover broad macroecological patterns

### 2021-07-15 manuscript/manuscript.tex
manuscript changes

%Title paper
\title{\vspace{-1cm}
Model linearity breeds contempt: using Bayesian non-linear models to uncover broad macroecological patterns}
% Model familiarity breeds contempt: simple models to compare properties of species' distributions
% Model simplicity breeds contempt: simple models to compare properties of species' distributions
% Model complexity breeds contempt: using non-linear models to compare basic properties of species' distributions
\author[1,*]{\normalsize Bernat Bramon Mora}
\author[2,3]{\normalsize Antoine Guisan}
\author[1]{\normalsize Jake M.\ Alexander}
\affil[1]{\footnotesize Institute of Integrative Biology, ETH Zürich, Zürich, Switzerland}
\affil[2]{\footnotesize Department of Ecology and Evolution, University of Lausanne, Lausanne, Switzerland}
\affil[2]{\footnotesize Department of Ecology and Evolution, University of Lausanne, Lausanne, Switzerland}
\affil[3]{\footnotesize Institute of Earth Surface Dynamics, University of Lausanne, Lausanne, Switzerland}
\affil[*]{\footnotesize  bernat.bramon@gmail.com}

@@

\section*{Abstract}
Species' realized niches are classically pictured as bell-shaped probability distributions. These distributions, however, can actually take many different forms. For example, fat-tailed or skewed responses are very common across fields, as these can naturally emerge as a result of several ecological processes. While one does not need to know the shape of species' distributions to effectively model them, studying their basic form can teach us a lot about the ways climatic processes and historical contingencies have shaped ecological communities. Unfortunately, we still lack a general understanding of the basic properties describing the shape of species' distributions, and much less is known about how these compare to each other across gradients. Here, we use a set of Bayesian non-linear models to uncover such properties. These models account for all prior knowledge we have regarding species' realized niches, including expert knowledge of their environmental preferences and ecological strategies. With this approach, we are able to distil the shape of empirical plant distributions, which helps us tackle long-standing hypotheses regarding the way ecological communities are assembled across space. In particular, we study the relationship between several properties of distributions, such as the link between species' range size and elevation, revealing the existence of broad macroecological patterns along environmental gradients. Moreover, we are able to shed light on the extent to which some aspects of the shape of observed realized niches---such as kurtosis and skewness of the distributions---could be intrinsic properties of species' historical contexts. Overall, our approach offers a useful statistical framework to understand the shape of species' distributions, and our results provide an unprecedented perspective of the way systems of many species are distributed along environmental gradients.

%Species' realized niches are often visualized as bell-shaped probability distributions. These, however, can take many different forms. For example, fat-tailed or skewed distributions are very common across fields, as these can naturally emerge as a result of several ecological processes. While one does not need to know the shape of species' distributions to effectively model them, there is a lot that can be learned about how climatic processes and historical contingencies differently affected species' distributions by studying the basic form. Indeed, we still lack a general understanding of the basic properties describing the shape of species' distributions, and much less is known about how these compare to each other across gradients. Here, we use a set of Bayesian non-linear models to uncover such properties. These models account for all prior knowledge we have regarding species' niches, including expert knowledge of their environmental preferences and ecological strategies. With this approach, we are able to study the shape empirical plant distributions, revealing general patterns regarding the way communities are assembled across space. In particular, by learning about the shape of distributions, we tackle long-standing hypothesis regarding broad macroecological patterns that are present along environmental gradients. We found conclusive evidence of the relationship between several properties of distributions, including the link between species' range size and elevation, or their general level of skewness. Perhaps most importantly, we are able to shed light on the extent to which some aspects of the shape of observed realized niches---such as kurtosis and skewness of the distributions---could be intrinsic properties of species' historical context. Overall, our approach offers a useful statistical framework to understand the shape of species' distributions, and our results provide an unprecedented perspective of the way systems of many species are distributed along environmental gradients.

%The shape of species' distributions along environmental gradients is part of old debates in ecology and biogeography. These have led the development of many statistical models centred around predicting the presence and absence of species over space and time. Useful to understand how populations of species will change along environmental gradients, they have become ecologists' compass to predict the effects of global climate change. However, there is a lot that can be learned about how climatic processes and historical contingencies have differently shaped species' distributions by studying instead their basic shape. Indeed, we still lack a general understanding of the shape of species' distributions, and much less is known about how these distributions compare to each other across gradients. Here, we use a set of Bayesian non-linear models to uncover the shape of species' realized niches. These models account for all prior knowledge we have regarding their shape, including expert knowledge of species' environmental preferences and ecological strategies. With this approach, we are able to uncover the general properties characterizing empirical plant distributions, revealing general patterns regarding the way communities are assembled. That is, by learning about the shape of distributions, we tackle long-standing hypothesis regarding broad macroecological patterns present along environmental gradients. In particular, we found conclusive evidence of the relationship between several properties of distributions, including the link between species' range size and elevation and their skewness along gradients. Finally, we are able to shed light on the extent to which some aspects of the shape of observed realized niches---such as kurtosis and skewness of the distributions---could be intrinsic properties of species. Overall, our approach offers a useful statistical framework to understand the shape of species' distributions, and our results provide an unprecedented perspective of the way systems of many species are distributed along environmental gradients.

%Species' distribution models have emerged as one of the most influential methodological advances in ecology and biogeography of the last decades. Useful to understand how populations of species will change along environmental gradients, they have become ecologists' compass to predict the effects of global climate change. That said, uncovering the mechanisms shaping species' realized niches has been one of the main driving forces behind the development of these models. That is, recent efforts have been often focused on understanding which biotic and abiotic factors are good predictors of species' niches---with an increasing effort placed in improving the predictive power of the statistical models. However, we still lack a general understanding of the shape of species' distributions, and much less is known about how these distributions compare to each other across gradients. Here, we use a set of Bayesian non-linear models to uncover the shape of species' realized niches. These models account for all prior knowledge we have regarding their shape, including expert knowledge of species' environmental preferences and physiology. With this approach, we are able to uncover the true shape of empirical species' distributions. Moreover, they allow us to tackle long-standing hypothesis regarding broad biogeographical patterns. In particular, we found conclusive evidence of the relationship between several properties of distributions, including the link between species' range size and elevation and their skewness along gradients. Finally, we are able to shed light on the extent to which some aspects of the shape of observed realized niches---such as kurtosis and skewness of the distributions---could be intrinsic properties of species. Overall, our approach offers a useful statistical framework to understand the shape of species' distributions, and our results provide an unprecedented perspective of the way systems of many species are distributed along environmental gradients.

@@

% REFERENCES TO LOOK AT FROM JAKE
% DATE EMAIL: 16.03.2021

%https://www.tandfonline.com/doi/abs/10.1080/01621459.1998.10474117
%https://hess.copernicus.org/preprints/hess-2015-24/
%https://agupubs.onlinelibrary.wiley.com/doi/epdf/10.1029/2009WR008933

%############# NEW INTRO ##############
%Bell shape is a good option, but it is unclear what the real shape is
One of the central goals of ecology is to understand the ways species are distributed across space and time. While many ecological textbooks assume the shape of species' realized niches to be unimodal and symmetric along environmental gradients \citep{krebsEcologyExperimentalAnalysis1972}, some have warned that empirical distributions can take many different forms \citep{austinModelsAnalysisSpecies1987, austinSpatialPredictionSpecies2002, sagarinMovingAssumptionsUnderstand2006}. In practice, there is a strong argument to be made in favour of assuming these to be bell shaped. Namely, if all that we are willing to assume about species' distributions is that these occupy finite geographic ranges, the most conservative statistical approach is to model their distribution as Gaussian (i.e.~the corresponding maximum entropy distribution; \citealt{frankCommonPatternsNature2009}). That said, there is currently no general agreement on the basic shape of species' realized niches. Indeed, many factors can play a role in defining their shape, and several natural processes can lead to non-normal distributions.

@@

%Skewed and fat-tails are common and they could well describe niches
Fat-tailed and skewed distributions are very common across fields. The former naturally emerges as a result of processes involving seasonality (e.g.~in communications patterns; \citealt{malmgrenPoissonianExplanationHeavy2008}) or some stochastic events (e.g.~in the spread of infectious diseases; \citealt{wongEvidenceThatCoronavirus2020}). Indeed, species' dispersal patterns have been shown to have fat tails due to the natural variability among individuals \citep{petrovskiiDispersalStatisticallyStructured2009}. This is important because one might expect environmental and individual variation to also be crucial factors determining the presence and absence of species along gradients, and fat tails are therefore a plausible property of species' realized niches. Similarly, several processes can lead to skewed distributions. For example, species might present asymmetric environmental tolerances along altitudinal gradients, allowing them to withstand different temperature extremes \citep{sundayGlobalAnalysisThermal2011}. Species might also experience abiotic and biotic pressures that increase or decrease along a temperature gradient, which could result in species' distributions presenting steeper declines towards warmer or colder environments \citep{normandImportanceAbioticStress2009}. Overall, many different properties could characterize species' realized niches, and every new shape entails different underlying hypotheses regarding the way communities are assembled over time \citep{damenSpatialPredictionsCommunity2017}.

%they can also reveal differences among species regarding the way they assemble along a gradient and broad macroecological patterns defining natural constraints shaping communities. 
%The general properties of species' distributions are valuable because they point to important ecological processes shaping species' realized niches. 
%For example, a relationship or lack thereof between range size and elevation would speak to classical ecological hypotheses (i.e.~Rapopor's rule;  \citealt{stevensElevationalGradientAltitudinal1992}) and could hint at the existence of general biogeographical constraints that shape species' distributions along gradients.
%Moreover, properties such as the skewness of species' distributions might capture other underlying processes regarding species' physiological tolerance to different environmental conditions \citep{kaufmanDiversityNewWorld1995}.
Comparing these properties across species allows us to study broad macroecological patterns that could be critical from a conservation and management perspective \citep{stevensElevationalGradientAltitudinal1992, channellDynamicBiogeographyConservation2000}. For example, the Rapoport's rule, a classsic biogeographical hypothesis, predicts species' ranges to increase with latitude or elevation \citep{stevensElevationalGradientAltitudinal1992}, hinting at the existence of general biogeographical constraints that shape species' distributions along gradients. This sort of macroecological patterns are interesting because they provide insights into the way different species assemble and establish in different environments \citep{linderGeographicRangeStructure2000}. That is, the differences in species' responses to the environment can shed light on how climatic processes and historical contingencies have differently shaped their distributions \citep{rohdeLatitudinalGradientsSpecies1992, helmuthBIOPHYSICSPHYSIOLOGICALECOLOGY2004, siefertHowClimateDispersal2015}. Uncovering the shape of species' realized niches and the extent to which these vary across species is nevertheless a challenging statistical problem to solve. Indeed, to this date, we do not have an effective way to parsimoniously compare the shape of the realized niches of many species along environmental gradients.
%Understanding the shape of species' realized niches and the extent to which these vary across species is therefore a crucial issue in ecology and biogeography (ref); however, we do not have an effective way to parsimoniously compare the realized niches of many species.
%Understanding the shape of species' realized niches and the extent to which these vary across species is however challenging statistical problem. That is, to this date, we do not have an effective way to parsimoniously compare the shape of the realized niches of many species.
%With all of that being said, measuring these properties is a challenging statistical problem, and we do not have an effective way to parsimoniously compare the shape of the realized niches of many species.

Over the last two decades, ecologists have developed a plethora of distribution models to try to untangle the factors that play a role in defining species' realized niches \citep{guisanPredictiveHabitatDistribution2000}. These models are fundamental to the scientific community for predicting changes in species' geographic distributions and the effects of environmental disturbances. Such frameworks, however, commonly assume an underlying linear relationship between covariates (but see `semiparametric models'; \citealt{norbergComprehensiveEvaluationPredictive2019}). This is useful because it simplifies the optimization process, but it might not be ideal when studying and comparing the shape of species' distributions along environmental gradients. First and foremost, a linear relationship between covariates often comes with a set of implicit mathematical constraints that might not be biologically justified. While this might not hinder the predictive performance of the models \citep{norbergComprehensiveEvaluationPredictive2019}, a direct biological interpretation of parameter estimates in linear models becomes increasingly difficult as one moves from unimodal and symmetric distributions \citep{terbraakWeightedAveragingLogistic1986, jamilGeneralizedLinearMixed2013} to fat-tailed or skewed responses \citep{huismanHierarchicalSetModels1993}. Second, the aforementioned structural constraints also limit our ability to include any prior information to our parameter estimates. Observations on species' geographic variation and optimal climatic conditions have long been documented, with extensive databases compiled by botanists and field ecologists documenting basic knowledge of species' realized niches (e.g.~\citealt{landoltFloraIndicativaOkologische2010}). That said, this information is rarely accounted for in most modelling approaches, likely because there is not a straightforward way to feed this information into the parameters of a linear model (\citealt{scherrerEcologicalIndicatorValues2019}; but see \citealt{terbraakWeightedAveragingLogistic1986, ovaskainenHowMakeMore2017}). Finally, some have proposed several non-linear structures to characterize several features of individual species' response curves \citep{huismanHierarchicalSetModels1993}. Setting aside the fact that the interpretation and comparison of parameter estimates becomes challenging following most of these model structures, these are generally not designed to jointly study different species, taking full advantage of modern statistical approaches (e.g. sharing information among species or accounting for parameter uncertainty; \citealt{evansProcessbasedRangeModeling2016}).

In this work, we rethink traditional modelling approaches and develop a conceptually simple---and yet statistical and computationally complex---statistical framework to revisit some classic hypothesis in ecology and biogeoraphy. In particular, we develop a Bayesian hierarchical model that accounts for all prior information that we have regarding the distribution of plant species along an elevation gradient in the Swiss Alps, including expert knowledge of species environmental indicator values, range sizes, and plant ecological strategies. We start by considering species' response curves as Gaussian distributed, and then we adapt our model to allow non-linear responses characterizing skewed and long-tailed distributions. Using this statistical framework, we are able to compare the basic properties of the realized niche of multiple species, testing for the existence of broad macroecological patterns. Comparing the posterior distribution of those parameters that control for the shape of distributions, we are also able to showcase variation in the way different types of species, such as native or neophytes, might respond to the environment. More generally, we are able to uncover the approximate shape of empirical plant distributions and answer fundamental questions regarding the way systems of many species are distributed along environmental gradients.

@@

\section*{Methods}
\subsection*{Empirical data}
We studied the distribution of plant communities along an elevation gradient. To do so, we combined two different datasets: i) one describing the co-occurrence of species across multiple open grasslands in the Swiss Alps \citep{randinClimateChangePlant2009}, and ii) an extensive floristic database containing environmental and physiological traits for all vegetation across Switzerland \citep{landoltFloraIndicativaOkologische2010}.

\subsubsection*{Distribution data}
We used data describing the distribution of 798 species across 912 sites covering most of the mountain region of the Western Alps in the Canton de Vaud (Switzerland; \citealt{scherrerEcologicalIndicatorValues2019}). Each of these sites is a $8\times 8\,\text{m}$ plot placed somewhere along an elevation range from $375\,\text{m}$ to $3210\,\text{m}$. In all sites, presence/absence data as well as Braun-Blanquet abundance-dominance classes were recorded for all species. Additionally, we used meteorological data provided by \citet{scherrerEcologicalIndicatorValues2019}, containing multiple variables characterizing the climate in each site at high spatial resolution ($25\,\text{m}$). This dataset was compiled based on 30 years (1961–1990) of records from national weather stations. Since most of the data is highly correlated, we calculated the main axes of variation of the following scaled variables: daily minimum, maximum and average temperature; sum of growing degree-days above $5^{\circ}\text{C}$; mean temperature of wettest quarter; annual precipitation, precipitation seasonality, and precipitation of driest quarter (Supplementary Fig.~1).  %Missing TabsY

\subsubsection*{Floristic data}
To complement the aforementioned distribution data, we used a floristic database of around 5500 vascular plants across Switzerland. Some of the information in this database has been previously shown to account for unexplained variation when used as explanatory variables in species' distribution models \citep{scherrerEcologicalIndicatorValues2019}.  It was built based on expert knowledge and phytosociological field experience of botanists and ecologists, and contains information regarding plants' environmental preferences and ecological strategies.

Species' environmental preferences in this database can be used to inform distribution models---e.g.~as an informative prior in a Bayesian framework. These are characterized following the ecological indicator values developed by \citet{landoltFloraIndicativaOkologische2010}, providing both an estimate of the average conditions in which a species can be found as well as a broad description of their range of variation. These values are provided for a range of 8 environmental variables, including temperature, continentality, light conditions, as well as moisture, acidity and nutrient content of the soil (see a full list and description of the ecological indicators in the Supplementary Table 1; \citealt{landoltFloraIndicativaOkologische2010}). In addition to species' environmental preferences, the floristic data also contains information on species introduction status (e.g.~identifying those species that are recent and historical range expanders) and change tendency (e.g.~indicating species that have shown decline or increase in their populations over the recent decades). We describe this information in more detail in the Supplementary Table 1.

@@

\subsection*{Baseline model}
There is a long list of model structures well suited to characterizing species' distributions (see \citealt{norbergComprehensiveEvaluationPredictive2019}). As a baseline model, however, we were interested in a hierarchical model that does not make any assumptions regarding the shape of the distributions, and yet explicitly incorporates all information that we have regarding plant's environmental preferences. More specifically, we wanted to account for the climatic indicator values and range of variation registered in the floristic database for all plants in our dataset. These two values provide basic information regarding plant's optimal environmental conditions and width of their distributions.

>>>>

%Title paper
\title{\vspace{-1cm}
Using Bayesian non-linear models to uncover broad macroecological patterns}
% Model familiarity breeds contempt: simple models to compare properties of species' distributions
% Model simplicity breeds contempt: simple models to compare properties of species' distributions
% Model complexity breeds contempt: using non-linear models to compare basic properties of species' distributions
\author[1,*]{\normalsize Bernat Bramon Mora}
\author[2,3]{\normalsize Antoine Guisan}
\author[1]{\normalsize Jake M.\ Alexander}
\affil[1]{\footnotesize Institute of Integrative Biology, ETH Zürich, Zürich, Switzerland}
\affil[2]{\footnotesize Department of Ecology and Evolution, University of Lausanne, Lausanne, Switzerland}
\affil[3]{\footnotesize Institute of Earth Surface Dynamics, University of Lausanne, Lausanne, Switzerland}
\affil[*]{\footnotesize  bernat.bramon@gmail.com}

@@

\section*{Abstract}
%Species' realized niches are classically pictured as bell-shaped probability distributions. These distributions, however, can actually take many different forms. For example, fat-tailed or skewed responses are very common in nature, as these can naturally emerge as a result of several ecological processes. While one does not need to know the shape of species' distributions to effectively model them, studying their basic form can teach us a lot about the ways climatic processes and historical contingencies shape ecological communities. Unfortunately, we still lack a general understanding of the basic properties describing the shape of species' distributions, and much less is known about how these compare to each other across gradients. Here, we use a set of Bayesian non-linear models to quantify such properties in empirical plant distributions. These models account for all prior knowledge we have regarding species' realized niches, including expert knowledge of their environmental preferences and ecological strategies. With this approach, we are able to distil the shape of plant distributions, tackling long-standing hypotheses regarding the way ecological communities are assembled across space. In particular, we study the relationship between several properties of distributions, such as the link between species' range size and elevation, revealing the existence of broad macroecological patterns along environmental gradients. We also shed light on the extent to which some aspects of the shape of observed realized niches---such as kurtosis and skewness of the distributions---could be intrinsic properties of species or the result of their historical contexts. Overall, our approach offers a useful statistical framework to understand the shape of species' distributions, and our results provide a novel perspective on the way systems of many species are distributed along environmental gradients.

% Am Nat abstract (200)
Species' realized niches are classically pictured as bell-shaped probability distributions. These distributions, however, can actually take many different forms, including fat-tailed or skewed responses. While one does not need to know the shape of species' distributions to effectively model them, studying their basic form can teach us a lot about the ways climatic processes and historical contingencies shape ecological communities. Unfortunately, we still lack a general understanding of the properties describing the shape of species' distributions, and how these compare to each other across gradients. Here, we use a set of Bayesian non-linear models to quantify such properties in empirical plant distributions. With this approach, we are able to distil the shape of plant distributions, tackling long-standing hypotheses regarding the way ecological communities are assembled across space. Studying the relationship between distribution properties, we revealed the existence of broad macroecological patterns along environmental gradients. We also shed light on the extent to which some aspects of the shape of observed realized niches---such as kurtosis and skewness of the distributions---could be intrinsic properties or the result of species’ historical contexts. Overall, our results provide a novel perspective on the way systems of many species are distributed along environmental gradients.

@@

One of the central goals of ecology is to understand the ways species are distributed across space and time. Many ecological textbooks assume the shape of species' realized niches to be unimodal and symmetric along environmental gradients \citep{krebsEcologyExperimentalAnalysis1972}. In practice, there is a strong argument to be made in favour of assuming these to be bell shaped. Namely, if all that we are willing to assume about species' distributions is that these occupy finite geographic ranges, the most conservative statistical approach is to model their distribution as Gaussian (i.e.~the corresponding maximum entropy distribution; \citealt{frankCommonPatternsNature2009}). However, some have warned that empirical distributions can take many different forms \citep{austinModelsAnalysisSpecies1987, austinSpatialPredictionSpecies2002}, and there is currently no general agreement on the basic shape of species' realized niches \citep{sagarinAbundantCentreDistribution2002, , sagarinMovingAssumptionsUnderstand2006}. Indeed, many factors can play a role in defining their shape, and several natural processes can lead to non-normal distributions.

@@

%Skewed and fat-tails are common and they could well describe niches
Fat-tailed and skewed distributions are very common across scientific fields. The former naturally emerges as a result of processes involving intermittent (e.g.~in email communications patterns; \citealt{malmgrenPoissonianExplanationHeavy2008}) or stochastic events (e.g.~in the spread of infectious diseases; \citealt{wongEvidenceThatCoronavirus2020}). Indeed, species' dispersal patterns have been shown to have fat tails due to the natural variability among individuals \citep{petrovskiiDispersalStatisticallyStructured2009}. This is important because one might expect environmental and individual variation to also be crucial factors determining the presence and absence of species along gradients, and fat tails are therefore a plausible property of species' realized niches. Similarly, several processes can lead to skewed distributions. For example, species might have asymmetric environmental tolerances along altitudinal gradients, allowing them to withstand different temperature extremes \citep{sundayGlobalAnalysisThermal2011}. Species might also experience abiotic and biotic pressures that increase or decrease along a temperature gradient, which could result in species' distributions presenting steeper declines towards warmer or colder environments \citep{normandImportanceAbioticStress2009}. Overall, many different properties could characterize species' realized niches, and every new shape entails different and potentially competing underlying hypotheses regarding the way communities are assembled over time \citep{damenSpatialPredictionsCommunity2017}.

%they can also reveal differences among species regarding the way they assemble along a gradient and broad macroecological patterns defining natural constraints shaping communities. 
%The general properties of species' distributions are valuable because they point to important ecological processes shaping species' realized niches. 
%For example, a relationship or lack thereof between range size and elevation would speak to classical ecological hypotheses (i.e.~Rapopor's rule;  \citealt{stevensElevationalGradientAltitudinal1992}) and could hint at the existence of general biogeographical constraints that shape species' distributions along gradients.
%Moreover, properties such as the skewness of species' distributions might capture other underlying processes regarding species' physiological tolerance to different environmental conditions \citep{kaufmanDiversityNewWorld1995}.
Comparing these properties across species allows us to study broad macroecological patterns that could be critical from a conservation and management perspective \citep{stevensElevationalGradientAltitudinal1992, channellDynamicBiogeographyConservation2000}. For example, Rapoport's rule, a classic biogeographical hypothesis, predicts species' range size to increase with latitude or elevation \citep{stevensElevationalGradientAltitudinal1992}, hinting at the existence of general biogeographical constraints that shape species' distributions along gradients. These sorts of macroecological patterns are interesting because they provide insights into the way different species assemble and establish in different environments \citep{linderGeographicRangeStructure2000}. That is, the differences in species' responses to the environment can shed light on how climatic processes and historical contingencies have differently shaped their distributions \citep{rohdeLatitudinalGradientsSpecies1992, helmuthBIOPHYSICSPHYSIOLOGICALECOLOGY2004, siefertHowClimateDispersal2015}. Uncovering the shape of species' realized niches and the extent to which these vary across species is nevertheless a challenging statistical problem. Indeed, to this date, we do not have an effective way to parsimoniously compare the shape of the realized niches of many species along environmental gradients.
%Understanding the shape of species' realized niches and the extent to which these vary across species is therefore a crucial issue in ecology and biogeography (ref); however, we do not have an effective way to parsimoniously compare the realized niches of many species.
%Understanding the shape of species' realized niches and the extent to which these vary across species is however challenging statistical problem. That is, to this date, we do not have an effective way to parsimoniously compare the shape of the realized niches of many species.
%With all of that being said, measuring these properties is a challenging statistical problem, and we do not have an effective way to parsimoniously compare the shape of the realized niches of many species.

Over the last two decades, ecologists have developed a plethora of distribution models to try to untangle the factors that play a role in defining species' realized niches \citep{guisanPredictiveHabitatDistribution2000, zurellEcologicalNicheModeling2019}. These models are fundamental to the scientific community for predicting changes in species' geographic distributions and the effects of environmental disturbances. Such frameworks, however, commonly assume an underlying linear relationship between covariates (but see `semiparametric models'; \citealt{norbergComprehensiveEvaluationPredictive2019}). This is useful because it simplifies the optimization process, but it might not be ideal when studying and comparing the shape of species' distributions along environmental gradients. First and foremost, a linear relationship between covariates often comes with a set of implicit mathematical constraints. These might not hinder the predictive performance of the models \citep{norbergComprehensiveEvaluationPredictive2019}, but a direct biological interpretation of parameter estimates in linear models becomes increasingly difficult as one moves from unimodal and symmetric distributions \citep{terbraakWeightedAveragingLogistic1986, jamilGeneralizedLinearMixed2013} to fat-tailed or skewed responses \citep{huismanHierarchicalSetModels1993}. Second, these mathematical constraints also limit our ability to include any prior information to our parameter estimates. Observations of species' geographic variation and optimal climatic conditions have long been documented, with extensive databases compiled by botanists and field ecologists documenting basic knowledge of species' realized niches (e.g.~\citealt{landoltFloraIndicativaOkologische2010}). That said, this information is rarely accounted for in most modelling approaches, likely because there is not a straightforward way to feed this information into the parameters of a linear model (\citealt{scherrerEcologicalIndicatorValues2019}; but see \citealt{terbraakWeightedAveragingLogistic1986, ovaskainenHowMakeMore2017}). Finally, some have proposed several non-linear structures to characterize several features of individual species' response curves \citep{huismanHierarchicalSetModels1993}. Setting aside the fact that the interpretation and comparison of parameter estimates becomes challenging following most of these model structures, these are generally not designed to jointly study different species, and therefore taking full advantage of modern statistical approaches (e.g. sharing information among species or accounting for parameter uncertainty; \citealt{evansProcessbasedRangeModeling2016}).

In this work, we develop a set of Bayesian hierarchical distribution models to study the shape of empirical plant distributions. We start by considering species' response curves as Gaussian distributed, and then we adapt our model to allow non-linear responses characterizing skewed and long-tailed distributions. This allows us to measure different properties of species' realized niches while accounting for all prior information regarding these distributions, including expert knowledge of species' environmental indicator values, range sizes, and plant ecological strategies. Using our statistical framework, we study the distribution of plant species along an elevation gradient in the Swiss Alps, revisiting some classic hypotheses in ecology and biogeoraphy. Specifically, we are able to compare the basic properties of the realized niche of multiple species, testing for the existence of broad macroecological patterns. Comparing the posterior distribution of those parameters that control for the shape of distributions, we are also able to showcase variation in the way different types of species, such as native or non-native, might respond to the environment. More generally, we are able to uncover the approximate shape of empirical plant distributions and answer fundamental questions regarding the way systems of many species are distributed along environmental gradients.

%In this work, we rethink traditional modelling approaches and develop a conceptually simple statistical framework to revisit some classic hypotheses in ecology and biogeoraphy. Specifically, we develop a Bayesian hierarchical model that accounts for all prior information that we have regarding the distribution of plant species along an elevation gradient in the Swiss Alps, including expert knowledge of species environmental indicator values, range sizes, and plant ecological strategies. We start by considering species' response curves as Gaussian distributed, and then we adapt our model to allow non-linear responses characterizing skewed and long-tailed distributions. Using this statistical framework, we are able to compare the basic properties of the realized niche of multiple species, testing for the existence of broad macroecological patterns. Comparing the posterior distribution of those parameters that control for the shape of distributions, we are also able to showcase variation in the way different types of species, such as native or non-native, might respond to the environment. More generally, we are able to uncover the approximate shape of empirical plant distributions and answer fundamental questions regarding the way systems of many species are distributed along environmental gradients.

@@

\section*{Methods}
\subsection*{Empirical data}
We studied the distribution of plant communities along an elevation gradient. To do so, we combined two different datasets: i) one describing the co-occurrence of species across multiple open grasslands in the Swiss Alps \citep{randinClimateChangePlant2009}, and ii) an extensive plant-attribute database containing environmental and life-history traits for all plant species across Switzerland \citep{landoltFloraIndicativaOkologische2010}.

\subsubsection*{Distribution data}
We used data describing the distribution of 798 species across 912 sites covering most of the mountain region of the Western Alps in the Canton de Vaud (Switzerland; \citealt{scherrerEcologicalIndicatorValues2019}). Each of these sites is a $8\times 8\,\text{m}$ plot placed somewhere along an elevation range from $375\,\text{m}$ to $3210\,\text{m}$. In all sites, presence/absence data as well as Braun-Blanquet abundance-dominance classes were recorded for all species. Additionally, we used meteorological data provided by \citet{scherrerEcologicalIndicatorValues2019}, containing multiple variables characterizing the climate in each site at high spatial resolution ($25\,\text{m}$). This dataset was compiled based on 30 years (1961–1990) of records from national weather stations. Since most of the data are highly correlated, we used a Principal Component Analysis to calculate the main axes of variation of the following scaled variables: daily minimum, maximum and average temperature; sum of growing degree-days above $5^{\circ}\text{C}$; mean temperature of wettest quarter; annual precipitation, precipitation seasonality, and precipitation of driest quarter (Supplementary Fig.~1). Notice that the first axis of variation is positively correlated with elevation and negatively correlated to temperature, while the second axis is positively correlation to precipitation seasonality (Supplementary Fig.~1). % TabsY

\subsubsection*{Plant attributes}
To complement the aforementioned distribution data, we used a attribute database of around 5500 vascular plants across Switzerland. Notice that some of the information in this database has been previously shown to account for unexplained variation when used as explanatory variables in species' distribution models \citep{scherrerEcologicalIndicatorValues2019}.  It was built based on expert knowledge and phytosociological field experience of botanists and ecologists, and contains information regarding plants' environmental preferences and ecological strategies.

Species' environmental preferences in this database can be used to inform distribution models---e.g.~as an informative prior in a Bayesian framework. These are characterized following the ecological indicator values developed by \citet{landoltFloraIndicativaOkologische2010}, providing both an estimate of the average conditions in which a species can be found as well as a broad description of their range of variation. These values are provided for a range of 8 environmental variables, including temperature, continentality, light conditions, as well as moisture, acidity and nutrient content of the soil (see a full list and description of the ecological indicators in the Supplementary Table 1; \citealt{landoltFloraIndicativaOkologische2010}). In addition to species' environmental preferences, the attribute data also contain information on species' introduction status (e.g.~identifying those species that are recent and historical range expanders) and change tendency (e.g.~indicating species that have shown decline or increase in their populations over the recent decades). We describe this information in more detail in Supplementary Table 1.

@@

\subsection*{Baseline model}
There is a long list of model structures well suited to characterizing species' distributions (see \citealt{norbergComprehensiveEvaluationPredictive2019}). As a baseline model, however, we were interested in a hierarchical model that does not make any assumptions regarding the shape of the distributions, and yet explicitly incorporates all information that we have regarding plant's environmental preferences. More specifically, we wanted to account for the climatic indicator values and range of variation registered in the attribute database for all plants in our dataset. These two values provide basic information regarding plants' optimal environmental conditions and the width of their distributions.

\subsubsection*{Response curve}
To choose an appropriate response curve, we first need to agree on what we truly know about the system. Given the prior information that we have about the system, we know that species occupy specific geographic ranges; therefore, we know that their distributions have finite variance. While we could also assume that many other factors might influence species' presence in a given site---e.g.~the biotic interactions among species in the site---we do not necessarily have an \textit{a priori} expectation of how exactly these factors will influence the shape of species' distributions. Therefore, for this baseline model, we choose the maximum entropy distribution with finite variance: a Gaussian distribution (Fig~\ref{fig:response}a). That is, given the presence/absence or abundance $y_{ij}$ of any species $i$ in any given site $j$, and an environmental variable $x_{j}$, we can define species' responses to the environment as
\begin{equation}
\begin{split}
y_{ij} & \sim F\left(p_{ij}\right) \\
\text{log}\left(p_{ij}\right) & = -\alpha_{i} - \gamma_{i} \left(x_{j}-\beta_{i}\right)^2 ,
\end{split}
\label{eq:baseline-response}
\end{equation}
where $F$ is the likelihood function, and $\alpha_i$, $\beta_i$, and $\gamma^{-1}_i$ describe the amplitude of the probability $p_{ij}$, species' average climatic suitability and range of variation along the environmental gradient, respectively. Notice that $F$ characterizes a Binomial distribution when considering binary data, and it characterizes an ordered categorical likelihood function when we consider Braun-Blanquet abundance-dominance classes as response variables (see the full description of both models in the Supplementary Methods). For the sake of simplicity, we use only one environmental variable to characterize the species' probability distribution. That said, this model can easily be generalized to account for multiple predictors (see Supplementary Methods).
\begin{figure}[ht]
  \centering
    \includegraphics[width=1\textwidth]{figures/all-distributions}
	   \caption{Different response curves. Panel (a) shows the distribution shapes characterized by Eq.~(\ref{eq:baseline-response}) for different values of $\beta$, $\alpha$ and $\gamma$. Panel (b) shows the distribution shapes characterized by Eq.~(\ref{eq:generalized-response}) for different values of $\beta$, $\alpha$ and $\nu$, when $\gamma=1$. Panel (c) shows the distribution shapes characterized by Eq.~(\ref{eq:skewed-response}) for different values of $\beta$, $\alpha$ and $\lambda$, when $\gamma=1$. Panel (d) shows the distribution shapes characterized by Eq.~(\ref{eq:heavy-skewed-response}) for different values of $\beta$, $\alpha$, $\lambda$ and $\nu$, when $\gamma=1$. Notice that for all panels, we chose $\alpha$ values such that $\text{p}_i = \exp(-\alpha_i)$.
	   }
      \label{fig:response}
\end{figure}
